---
title: "CISER Statistics in R Workshop: Day 1"
subtitle: "Presented by the Center for Interdisciplinary Statistical Education and Research"
output: 
  slidy_presentation: default

---

Workshop Overview
======================================================== 

```{r init, include = F}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, error = F)
```

```{r load, echo = F, include = F}
source("../../../classes/stat511/R/show_png.R")
library(tidyverse)
```

* Day 1
    + Sample Size Calculations
    + Using R Data Frames 
    + t-tests
    + Linear regression

* Day 2
    + Reading data into R  
    + Fixed Effects ANOVAs
    + Random Effects Models


Sample Size Calculations in R
======================================================== 

<!-- INSERT PHOTO OF DOG BLOOD DRAW -->
* Study to compare different solutions (A and B) for platelet storage in dogs
* Primary interest is to compare clumping of A to clumping of B
* Plan to analyze results using a t-test
* HOW DO I FIND OUT HOW TO ESTIMATE SAMPLE SIZES IN R?


Extending the Functionality of R
======================================================== 

* An R package is a collection of R functions, data, and compiled code in a well defined format
* Packages extend the base functionality of R  

```{r, echo = F}
show_png("../../../classes/stat511/img/img-000139.png", file = T)
```


Finding a Package for Sample Size Calculations
======================================================== 
* Typically try:
    + CRAN Task Views: https://cran.r-project.org/web/views/
    + Google Search: "R power calculation"

* The `pwr` package in R can be used to make power calculations for some simple cases including
    + Tests for one or two normal populations
    + Tests for one or two proportions
* A package can be installed using the `install.packages()` function

```{r, eval = F}
install.packages('pwr')
```

* Every time that you restart R and want to use the `pwr` package, run the following to make the functions available

```{r}
library(pwr)
```

* The `pwr.t.test` function can be used to make power calculations for 1-sample, 2-sample and paired t-tests   


    
Sample Size Calculations in R
========================================================     
* To make a sample size calculation, we need
    + alpha level = rate of being wrong under the null
    + power = rate of being right given model assumptions at the minimal significant difference
    + error standard deviation = expected deviation from treatment mean
    + minimal significant difference = the smallest difference we still want to be able to detect

* We want to make a power calculation for a one-sample t-test
* Assume this information is known
    + Type I error rate ($\alpha$) = 0.05 
    + Type II error rate ($\beta$) = 0.20
    + error standard deviation = 1.25
    + difference in means = 0.6
    
`pwr` package: t-test power calculations
======================================================== 

* Requires entering
    + Effect Size: `d` =  difference in means divided by estimated standard deviation
    + Type I Error Rate: `sig.level`
    + Power: `power` = 1 - Type II Error Rate
    + Type of t-test: `type` = `two.sample`, `one.sample`, `paired`
    + Alternative: `two.sided`, `greater`, `less`

```{r, eval = F}
library(pwr)
pwr.t.test(d = 0.6 / 1.25, sig.level = 0.05, power = 0.80, type = "one.sample", alternative = "two.sided")
```


`pwr` package: one-sample t-test power calculation
======================================================== 

```{r, echo = F}
show_png("../../../classes/stat511/img/img-000138.png", file = T)
```


```{r}
library(pwr)
pwr.t.test(d = 0.6 / 1.25, sig.level = 0.05, power = 0.80, type = "one.sample", alternative = "two.sided")
```


Exercise 1: Sample size for paired t-test
======================================================== 
* Assume that a sample of cats are to undergo two separate treatments (trt A and trt B) for pain in randomized order
* Assume:
    + We plan on testing for differences between trt A and trt B with a paired t-test (equivalent to one-sample t-test on differences in pain)
    + Want to be able to detect an average difference in treatments of 0.75 or greater
    + Standard deviation of differences is 1.5
    + Want a power of 0.90 at an alpha value of 0.05



Exercise 1 Solution
======================================================== 

```{r}
pwr.t.test(d = 0.75 / 1.5, sig.level = 0.05, power = 0.90, type = "one.sample", alternative = "two.sided")
```

Tooth Growth Dataset
======================================================== 
ToothGrowth data set contains the result from an experiment studying the effect of vitamin C on tooth growth in 60 Guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

* The `head` function returns the first or last parts of a vector, matrix, table, data frame or function
* The `head` function is useful in getting a quick sense of what your data look like

```{r}
head(ToothGrowth)
```



data.frames for working with data of multiple types in R
========================================================

* data.frames allow for row / column subsetting of multi-class data
* ToothGrowth is a data.frame
* We can verify this using the `class` function

```{r}
class(ToothGrowth)
```

* The class function is very helpful in identifying whether variables are of the appropriate class before we conduct an analysis
* We can check the class of variable in the ToothGrowth data.frame 

```{r}
class(ToothGrowth$len)
```

* Placing the `$` after a data.frame will allow us to access particular variables in the data.frame

```{r}
ToothGrowth$len
```

Exercise 2: Determine classes of variables
======================================================== 
* Determine the class of `supp` and `dose` from the ToothGrowth data


Exercise 2 Solution
========================================================

* The class of the `supp` variable is a factor and the class of the `dose` variable is numeric
* Factors are also called categorical variables

```{r}
class(ToothGrowth$supp)
class(ToothGrowth$dose)
```

Creating simple data.frames directly in R
========================================================

```{r}
df <- data.frame(
  a = c(1, 2, 3),
  b = c("x", "y", "z"),
  c = c(TRUE, TRUE, FALSE),
  d = factor(c("cat", "dog", "horse"))
  )
```

Taking subsets of rows / columns
========================================================

```{r}
# Extracting a single row
df[3, ]

# Extracting a single column
df[, 2]

# Extracting multiple rows
df[c(1,3), ]

# Extracting multiple columns
df[, c(2,3)]

# Removing rows
df[-c(2,3), ]

# Removing columns
df[, -3]

# Extracting a column by name
df[, c("a", "c")]
df$a

# Returning rows that meeting only a certain criteria
df[df$c == TRUE & df$a >1, ]
```

Exercise 3: Subsetting the ToothGrowth data
======================================================== 
* Imagine we want to compare tooth growth for vitamin C doses of 0.5 and 2 mg per day with a delivery method of orange juice ignoring all of the other collected data

* Extract the lengths separately for the orange juice 0.5 mg/day group and the orange juice 2.0mg/day group

Exercise 3 Solution
======================================================== 

```{r}
len05 <- ToothGrowth[ToothGrowth$dose == 0.5 & ToothGrowth$supp == "OJ",]$len
len20 <- ToothGrowth[ToothGrowth$dose == 2.0 & ToothGrowth$supp == "OJ",]$len
```

t-test for the ToothGrowth Data
======================================================== 

* A t-test can be run in R using the function `t.test`
* For instance, lets say that we wanted to run a t-test to test for a difference between the 0.5 mg/day and the 2.0mg/day orange juice groups from the ToothGrowth data.

```{r}
# Conduct a two sample, two-sided Welch's t-test
t.test(len20, len05)
```

* Running the `t.test` function prints out a few helpful pieces of information including
    + The t-statistic
    + The error degrees of freedom
    + The p-value
    + 95 \% confidence intervals on the estimated difference in means
    + Sample means for the two groups


Options for the `t.test` function
========================================================

* We can browse the `t.test` options by typing

```{r}
?t.test
```

* Some of the most useful options are:
    + `alternative`: Run a 1-sided t-test
    + `paired`: Conducting paired t-tests
    + `var.equal`: Conducting equal variance t-tests
    
* To conduct a one sample t-test to test the null hypothesis that the mean length for the 0.5 mg/day group was >= 15

```{r}
t.test(len05, mu = 15, alternative = "less")

```

Exercise 4: T-tests
======================================================== 

* Conduct a one-sided t-test to test the null hypothesis that the mean length for the 2.0 mg/day group is <= 24



Exercise 4 Solution
======================================================== 

```{r}
t.test(len20, mu = 24, alternative = "greater")

```


