---
title: "CISER Statistics in R Workshop: Day 2"
subtitle: "Presented by the Center for Interdisciplinary Statistical Education and Research"
output: 
  slidy_presentation: default

---

Workshop Overview
======================================================== 

```{r init, include = F}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, error = F)
```

```{r load, echo = F, include = F}
source("../../../classes/stat511/R/show_png.R")
library(tidyverse)
```

* Day 1
    + Sample Size Calculations
    + Using R Data Frames 
    + t-tests

* Day 2
    + Linear regression
    + Full/reduced drop tests 
    + Reading data into R


Motor Trend Car Road Tests
======================================================== 

The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).


* mpg	 Miles/(US) gallon
* wt	 Weight (1000 lbs)
* am	 Transmission (0 = automatic, 1 = manual)
* gear	 Number of forward gears

```{r}
head(mtcars)
```

Simple Linear Regression in R
======================================================== 

* We can use the `mtcars` data to get a sense of the relationship between a car's weight and it's gas mileage

* The `plot` function can be used to create a scatter plot for continuous data

```{r}
plot(mtcars$wt, mtcars$mpg)
```

* The plot appears to show a relationship between the weight of a car and the gas mileage


Exercise 1: Construct a scatter plot
======================================================== 

* Construct a scatter plot of the miles per gallon vs. displacement

Exercise 1: Solution
======================================================== 

```{r}
plot(mtcars$disp, mtcars$wt)
```

Linear regression to understand the relationship between gas mileage and weight
======================================================== 
* We can use linear regression to model this data

* The function `lm` will fit a linear regression 

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
```

* The expression `mpg ~ wt` is known as a model formula, and is a key aspect of fitting various models in R

* The model formula can be read as: *mpg is modeled as a (linear) function of weight*

* We can look at a simple summary of the linear regression by typing

```{r}
summary(mod)
```

* Key metrics to look at are:
    + Parameter estimates (intercept and slope)
    + Standard error in the estimates
    + t-statistic for testing the hypothesis that each parameter is equal to zero
    + Adjusted R-squared indicates the expected out of sample variance explained for the same population
    + Overall test of significance for all parameters in the model except the intercept

Exercise 2: Fitting a linear model 
======================================================== 

* Fit a linear model of gross horsepower (hp) vs. displacement
* Compare this fit to a fit of gross horsepower (hp) vs. number of cylinders
* Compare the R-squared adjusted to get a sense of which variable better explains horsepower

Exercise 2: Solution
======================================================== 

* The R-squared adjusted for the model with displacement is 0.61
* The R-squared adjusted for the model with cyl is 0.68
* The R-squared adjusted is larger for the model with cyl

```{r}
mod <- lm(hp ~ disp, data = mtcars)
summary(mod)$adj.r.squared
mod <- lm(hp ~ cyl, data = mtcars)
summary(mod)$adj.r.squared
```

Extracting Statistics from a Fitted Regression Model
======================================================== 

* A fitted regression model comes with various statistics
* Depending on the type of regression model, there may be different statistics available
* You can check what is available by looking at the `value` section in the help for the function that has been called
* Alternatively, you can look at the names of the list that was returned

```{r}
mod <- lm(hp ~ cyl, data = mtcars)
names(summary(mod))
summary(mod)$adj.r.squared
```

Fitting Models with Factor Variables
======================================================== 

* Lets consider how the mpg relates to the number of forward gears
* For this dataset, there are cars with 3, 4, and 5 total forward gears
* We could treat gears as a continuous variable and fit a linear model
* Below is a plot showing mpg vs. gear along with the best fit linear regression model with gear treated as a linear continuous predictor

```{r}
mod <- lm(mpg ~ gear, data = mtcars)
summary(mod)$adj.r.squared
plot(mtcars$gear, mtcars$mpg)
lines(mtcars$gear, fitted(mod))
```

* Assuming linearity of gear does not seem to adequately reflect the relationship between the number of gears and the gas mileage

Fitting Models with Factor Variables
======================================================== 
* We can treat gear as a factor variable
* We can include the factor function directly in the model formula

```{r}
mtcars$fgear <- factor(mtcars$gear)
mod <- lm(mpg ~ fgear, data = mtcars)
summary(mod)$adj.r.squared
plot(mtcars$gear, mtcars$mpg)
lines(mtcars$gear, fitted(mod))
lines(mtcars$gear, fitted(mod), col = "red")
```

* Treating gear as a categorical variable allows for more flexibility in the relationship
* The R-squared adjusted for mpg~factor(gear) is 0.39
* The R-squared adjusted for mpg~gear is 0.21

Multiple Regression with Interactions
======================================================== 

* We can construct multiple regression models by including more variables in the model
* If we are interested in making a model that best explains the variation in gas mileage, we might try something like

```{r}
mod <- lm(mpg ~ fgear + wt + carb + wt:carb, data = mtcars)
```

* In this model, we have included linear terms for wt, fgear, carb and an interaction between wt and carb
* Model terms (wt, fgear, carb, wt:carb) are included additively using the `+` operator
* Interaction terms can be created using the `:` operator
* This model formula can be simplified by instead using the `*` operator
* The `*` operator includes the interaction and all lower order terms

```{r}
mod <- lm(mpg ~ fgear + wt*carb, data = mtcars)
summary(mod)$adj.r.squared
```

Exercise 3: Fit a multiple regression model
======================================================== 

* Fit a multiple regression model where you explain variation in horsepower (hp) using the variables displacement (disp), automatic v. manual (am) and the interaction
* Is there any evidence to suggest an interaction between cyl and wt?

Exercise 3: Solution
======================================================== 

* There appears to be substantial evidence against the null hypothesis of no interaction between displacement and whether the car is automatic/manual (t = 3.85, df = 28, p = 0.0006).

```{r}
mod <- lm(hp ~ disp + am + disp:am, data = mtcars)
summary(mod)
```

Full / Reduced Drop Tests for Testing Categorical Variables
======================================================== 

* Consider the variable we constructed called `fgear`
* The model formula `mpg ~ fgear + wt` allows us to 

```{r}
mod <- lm(mpg ~ wt + fgear, data = mtcars)
summary(mod)
```

* `summary` does not provide overall evidence concerning a `gear` effect
* We can drop terms and summarise the evidence by comparing full and reduced models

```{r}
mod0 <- lm(mpg ~ wt, data = mtcars)
anova(mod, mod0)
```

* This will provide us with a p-value that summarises the effect of `fgear` conditional on `wt` being in the model

* We can obtain on test for each term in the model using the `drop1` command

```{r}
drop1(mod, test = "F")
```


Exercise 5: Test multiple categorical variables
======================================================== 

* Using the `mtcars` data, fit a model with `mpg` as the response, and `cyl` and `gear` as **categorical** explanatory variables
* Obtain F drop test statistics for each of the two factors (cyl, gear) conditional on the other being in the model

Exercise 5: Solution
======================================================== 

```{r}
mtcars$fcyl <- factor(mtcars$cyl)
drop1(lm(mpg ~ fgear + fcyl, data = mtcars))
```


Reading your data into R
======================================================== 

* Data can be read into R in a variety of formats
* Here we focus on tabular formats where data is stored in basic text files
* We also assume that each column is separated with a particular separator 
* Examples include tab delimited files and comma separated files
* We can read in a file with comma separated values (csv) using

```{r reading_data, eval = F}
platelet <- read.table("platelet.csv", sep = ",", header = T)
```


File Access / Working Directory
======================================================== 
* The working directory is the folder that you are working in 
* When accessing files in working directory, you do not need to provide a file path
* `getwd()` tells you your current working directory
* If the file you want to access is not in the current working directory, you can change the working directory with `setwd()`

```{r, eval = F}
setwd("C:/mystuff/") 
```

* The backslash "\\" is an escape character in R. If you use it for file paths, you must use it twice
```{r, eval = F}
setwd("C:\\mystuff\\")
```

* Alternatively, you can use the dropdown for changing the working directory


Exercise 5: Read data into R from the file `catpain.csv`
======================================================== 
* Try reading in the file `catpain.csv` 


Exercise 5: Solution
======================================================== 

* Your solution will differ slightly from this
```{r, eval = F}
setwd("./data")
catpain <- read.table("catpain.csv", header = T, sep = ",")
```

